---
title: "prsperf"
output: html_document
editor_options: 
  chunk_output_type: console
---

<!-- 
Run this 'development' chunk

Store every call to library() that you need to run chunks line by line, as in a classical Rmd for analysis
-->

```{r development, include=FALSE}
library(testthat)
library(devtools)
```

<!--
# Description of your package

This will fill the description of your package.
Fill and run the content of this chunk, before anything else. 

Note: when you will use other flat templates, this part will be in a separate file. Do not be surprised!
--> 

```{r description, eval=FALSE}
# Describe your package
fusen::fill_description(
  pkg = here::here(),
  fields = list(
    Title = "Estimate Theoretical Performance of PRS",
    Description = "A set of functions designed to estimate the optimal performance of Polygenic Risk Scores based on heritablity and population prevalence. The functions are obtained from: https://doi.org/10.1186/1471-2164-13-340",
    `Authors@R` = c(
      person("Michael", "Levin", email = "michael.levin@pennmedicine.upenn.edu", role = c("aut", "cre"), comment = c(ORCID = "0000-0002-9937-9932"))
    )
  )
)
# Define License with use_*_license()
usethis::use_mit_license("Michael Levin")
```

# Estimate Maximal AUC

This function allows the user to estimate the maximal AUC based on population prevalence (`K`) and proportion of variance explained (`pve`).

```{r function-max_auc}
#' Estimate the Maximum AUC
#' 
#' Estimate the maximum AUC of a polygenic risk score based on the population prevalence of a disease `k` and the proportion of variability explained by the PRS `pve`.
#' 
#' @param k (numeric; range 0-1) Population prevalence of the disease of interest
#' @param pve (numeric; range 0-1) Proportion of variability explained by the PRS
#' @param n.bins (integer) Number of bins for estimating the AUC (default = 100)
#'
#' @return A numeric value for the maximum AUC
#' @export
max_auc <- function(k, pve, n.bins=100){
    checkmate::assert_numeric(k, lower = 0, upper = 1)
    checkmate::assert_numeric(pve, lower = 0, upper = 1)
    checkmate::assert_numeric(n.bins)
  
    Q <- makeQ(n.bins=n.bins)
    #check concavity to ensure that Q is negative semidefinite
    eigen.vals <- eigen(x=Q, symmetric=TRUE, only.values=TRUE)$values
    if (!all(eigen.vals<=0)){ warning('The matrix Q is not concave') }
    #constraints: prevalence, pve, 0<=p, p<=1, sum(p)<=1
    Amat <- rbind((1:n.bins)/n.bins, (1:n.bins)^2/(n.bins^2), diag(n.bins), -diag(n.bins), rep(-1, n.bins))
    rhs <- c(k, k*(1-k)*pve+k^2, numeric(n.bins), rep(-1, n.bins), -1)
    #run quadratic program over p
    qp.res <- quadprog::solve.QP(Dmat=-Q, dvec=numeric(n.bins), Amat=t(Amat), bvec=rhs, meq=2)
    auc <- (-2*qp.res$value+n.bins^2*k)/(n.bins^2*k*(1-k))
    return(auc)
}
```

Here, we can use the example of type II diabetes provided in the original manuscript:
```{r example-max_auc}
max_auc(k = 0.13, pve = 0.26)
```
  
```{r tests-max_auc}
test_that("max_auc works", {
  res <- max_auc(k = 0.13, pve = 0.26)
  expect_true(inherits(max_auc, "function"))
  expect_type(res, "double")
})
```
  
  
```{r function-makeQ}
#' makeQ
#'
#' @param n.bins (integer) Number of bins
#'
#' @return a matrix

makeQ <- function(n.bins){
    mat <- matrix(0,nrow=n.bins,ncol=n.bins)
    for (i in 1:nrow(mat)){
        for (j in 1:ncol(mat)){
            if (i>j){ mat[i,j] <- -(n.bins+i)*j/2 } else { mat[i,j] <- -(n.bins+j)*i/2 }
        }
    }
    return(mat)
}
```


# Estimate optimal sensitivity

This function estimates the optimal (maximal or minimal) sensitivity of a PRS

```{r function-optSe}
#' Optimal Sensitivity
#'
#' These functions estimate the optimal (maximal or minimal) sensitivity of a PRS.
#' 
#' @param thresh.vector (numeric) Vector of thresholds
#' @param k (numeric; range 0-1) Population prevalence of the disease of interest
#' @param pve (numeric; range 0-1) Proportion of variability explained by the PRS
#' @param sp (numeric; range 0-1) Specificity
#' @param n.bins (integer) Number of bins for estimating the AUC (default = 100)
#' @param direction (character) `"max"` or `"min"`
#'
#' @return A numeric valuve for the optimal sensitivity
#' 
#' @export

opt_se <- function(k, pve, sp, n.bins=1000, thresh.vector=seq(from=1,to=n.bins,by=10), direction="max"){
    #vector of sensitivity at each threshold
    se.tmp.v <- apply(X=as.matrix(thresh.vector), MARGIN=1, FUN=optSeGivenThresh, k=k, pve=pve, n.bins=n.bins,
    sp=sp, direction=direction)
    se <- max(se.tmp.v)
    return(se)
}

```

```{r example-opt_se}
opt_se(k = 0.13, pve = 0.26, sp = 0.99)
```
  
```{r tests-opt_se}
test_that("opt_se works", {
  res <- opt_se(k = 0.13, pve = 0.26, sp = 0.99)
  expect_true(inherits(opt_se, "function")) 
  expect_type(res, "double")
})
```

```{r function-optSeGivenThresh}
optSeGivenThresh <- function(thresh, k, pve, sp, n.bins, direction="max"){

    obj <- c(rep(0, thresh), thresh:n.bins)/(n.bins*k)
    sp.coeff <- c(n.bins-0:(thresh-1), rep(0, n.bins-thresh+1))/(n.bins*(1-k))
    #constraints: avg risk=k, pve, sp, sum(p)=1, 0<=p, p<=1
    Amat <- rbind((0:n.bins)/n.bins, (0:n.bins)^2/(n.bins^2), sp.coeff, rep(1,n.bins+1), diag(n.bins+1), -diag(n.bins+1))
    rhs <- c(k, k*(1-k)*pve+k^2, sp, 1, numeric(n.bins+1), rep(-1,n.bins+1))
    lp.res <- lpSolve::lp(direction=direction, objective.in=obj, const.mat=Amat, const.dir=c(rep("=",4), 
    rep(">=", 2*n.bins+2)), const.rhs=rhs)
    return(lp.res$objval)
}

```

  
# Estimate optimal specificity
    
```{r function-opt_sp}
#' Optimal Specificity
#' 
#' These functions estimate the optimal (maximal or minimal) sensitivity of a PRS.
#' 
#' @param thresh.vector (numeric) Vector of thresholds
#' @param k (numeric; range 0-1) Population prevalence of the disease of interest
#' @param pve (numeric; range 0-1) Proportion of variability explained by the PRS
#' @param se (numeric; range 0-1) Sensitivity
#' @param n.bins (integer) Number of bins for estimating the AUC (default = 100)
#' @param direction (character) `"max"` or `"min"`
#'
#' @return A numeric valuve for the optimal sensitivity
#' 
#' @export
#' 
opt_sp <- function(k, pve, se, n.bins=1000, thresh.vector=seq(from=1,to=n.bins,by=10), direction="max"){
    checkmate::assert_numeric(k, lower = 0, upper = 1)
    checkmate::assert_numeric(pve, lower = 0, upper = 1)
    checkmate::assert_numeric(n.bins)
    checkmate::assert_numeric(se, lower = 0, upper = 1)
    checkmate::assert_choice(direction, choices = c("min", "max"))
    #vector of specificity at each threshold
    sp.tmp.v <- apply(X=as.matrix(thresh.vector), MARGIN=1, FUN=optSpGivenThresh, k=k, pve=pve, n.bins=n.bins,
    se=se, direction=direction)
    sp <- max(sp.tmp.v)
    return(sp)
}
```
  
```{r example-opt_sp}
opt_sp(k = 0.13, pve = 0.26, se = 0.99)
```
  
```{r tests-opt_sp}
test_that("opt_sp works", {
  res <- opt_sp(k = 0.13, pve = 0.26, se = 0.99)
  expect_true(inherits(opt_sp, "function")) 
  expect_type(res, "double")
})
```
  

```{r function-optSpGivenThresh}
optSpGivenThresh <- function(thresh, k, pve, se, n.bins, direction="max"){
    obj <- c(n.bins-0:(thresh-1), rep(0, n.bins-thresh+1))/(n.bins*(1-k))
    se.coeff <- c(rep(0, thresh), thresh:n.bins)/(n.bins*k)
    #constraints: avg risk=k, pve, sp, sum(p)=1, 0<=p, p<=1
    Amat <- rbind((0:n.bins)/n.bins, (0:n.bins)^2/(n.bins^2), se.coeff, rep(1,n.bins+1), diag(n.bins+1), -diag(n.bins+1))
    rhs <- c(k, k*(1-k)*pve+k^2, se, 1, numeric(n.bins+1), rep(-1,n.bins+1))
    lp.res <- lpSolve::lp(direction=direction, objective.in=obj, const.mat=Amat, const.dir=c(rep("=",4), 
    rep(">=", 2*n.bins+2)), const.rhs=rhs)
    return(lp.res$objval)
}
```


<!-- 
# Inflate your package

You're one inflate from paper to box.
Build your package from this very Rmd using `fusen::inflate()` 
-->


```{r development-inflate, eval=FALSE}
# Execute in the console directly
fusen::inflate(flat_file = "dev/auc.Rmd")
```

<!-- 
- Verify your `"DESCRIPTION"` file has been updated
- Verify your function is in `"R/"` directory
- Verify your test is in `"tests/testthat/"` directory
- Verify this Rmd appears in `"vignettes/"` directory 
-->
